Supporting Information:\\ NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-12-19 Manuscript/manuscriptPSsuppl.tex

\usepackage{upgreek}				
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{xr}
\externaldocument{manuscriptPS}

@@

\noindent
{\it POPS bilayers.} \\
The data in Table \ref{PSsystems} and Refs \citenum{macrogPOPS298Kcorrect,macrogPOPS298KwithK}
\todo{T. Piggot and M. Javanainen, please write simulation details} \\

\noindent
{\it POPC:POPS (5:1) mixtures with additional potassium ions} \\
The data in Table \ref{mixedIONsystems} and Refs \citenum{POPCpopsMACROG,POPCpopsMACROGwithK}
\todo{M. Javanainen, please write simulation details} \\

>>>>

\usepackage{upgreek}				
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{xr}
\externaldocument{manuscriptPS}
\usepackage{soul} % for highlights with \hl{}

@@

The POPC bilayer consisting of a total of 128 lipids was constructed from single lipid structure taken 
from Ref.~\cite{10.1016/j.dib.2015.09.013}. The bilayer was hydrated by a total of 5120 water 
molecules, corresponding to 40 water molecules per lipid.

The MacRog lipid parameters \cite{10.1016/j.chemphyslip.2015.07.002}, obtained from 
\cite{10.1016/j.dib.2015.09.013}, were used, along with the TIP3P water model 
\cite{10.1063/1.445869}. The simulation parameters were identical to those used for 
POPC/POPC mixtures with the MacRog force field, see below. The simulation was run for 200~ns, 
out of which \hl{??~ns: (this is not tabulated in the MS or the SI)} were used for analyses.

\noindent
{\it POPS bilayers.} \\
The data in Table \ref{PSsystems} and Refs \citenum{macrogPOPS298Kcorrect,macrogPOPS298KwithK}
\todo{T. Piggot and M. Javanainen, please write simulation details} \\
\hl{Piggot: please write about system with Na$^+$ counterions}.

This membrane was further simulated after replacing Na$^+$ counterions by K$^+$ ones, which 
were described by the \AA{}qvist parameters \cite{10.1021/j100384a009}. Other topologies
were unchanged. The simulation parameters were the same as used for the POPC/POPS mixtures with 
the MacRog force field, see below. The simulation was run for 200~ns, out of which 150~ns was used 
in the analyses.

\noindent
{\it POPC:POPS (5:1) mixtures with additional potassium ions} \\
The data in Table \ref{mixedIONsystems} and Refs \citenum{POPCpopsMACROG,POPCpopsMACROGwithK}

The bilayers containing a total of 120 POPC and 24 POPS molecules distributed evenly among the 
two leaflets were set up using CHARMM-GUI \cite{10.1002/jcc.20945}.
The bilayers were solvated by 5760 water molecules, corresponding to 40 water molecules for lipid.
Additionally, 24 potassium ions were added to neutralize the charge of the POPS head groups.

These bilayers were simulated in the presence of counterions only, as well as together with different
concentrations of CaCl$_2$ or KCl. For the former, concentrations of 100~mM, 300~mM, 1~M, and 3~M 
were considered, which corresponded to the amounts of 10/20, 31/62, 104/208, and 311/622 
of Ca$^{2+}$/Cl$^-$ ions. For membranes with KCl, concentrations of 500~mM, 1~M, 2~M, and 3~M of 
were considered, corresponding to 52, 104, 208, or 311 pairs of K$^+$ and Cl$^-$ ions.

The lipids were described using the MacRog model \cite{10.1021/jp5016627,10.1016/j.chemphyslip.2015.07.002,10.1016/j.dib.2016.03.067}. The POPC topologies were obtained from 
Ref.~\citenum{10.1016/j.dib.2015.09.013}, whereas the POPS topology was adapted from 
Ref.~\citenum{10.1016/j.dib.2016.03.067}. Notably, the chains were reversed in order to modify 
OPPS into POPS. Unlike the D stereoisomer present in Ref.~\cite{10.1016/j.dib.2016.03.067}, the 
POPS head group conformations generated by CHARMM-GUI are in the naturally occurring L stereoisomer.
TIP3P model \cite{10.1063/1.445869} was used for water, and the \AA{}qvist parameters 
\cite{10.1021/j100384a009}, standard for the OPLS force field, were employed for the ions.

The simulation parameters were taken from Ref.~\citenum{10.1016/j.dib.2016.03.067}. 
Periodic boundary conditions were employed in all dimensions. The list of neighbors was updated every step 
\hl{I used nstlist=1 because it's used in the mdp in the article quoted. In DOI: 10.1021/jp5016627, nstlist=10 is used}
The Lennard-Jones interactions were cut off at 1~nm, whereas the smooth particle mesh Ewald algorithm 
\cite{10.1063/1.470117} was used for electrostatics. Dispersion correction \cite{10.1021/jp0735987} 
was applied to both energy and pressure. Temperature was kept constant at 298~K using the 
Nos\'{e}--Hoover thermostat \cite{10.1063/1.447334,10.1103/PhysRevA.31.1695}. The lipids and 
the solvent were coupled separately, and a time constant of 0.4~ps was employed. Pressure was 
maintained semi-isotropically at 1~bar using the Parrinello--Rahman barostat with a time constant of 1~ps 
\cite{10.1063/1.328693}. All non-water bonds were constrained with the LINCS algorithm 
\cite{10.1002/(SICI)1096-987X(199709)18:12<1463::AID-JCC4>3.0.CO;2-H,10.1021/ct700200b},
whereas the bonds in water molecules were constrained using SETTLE \cite{10.1002/jcc.540130805}.
The integration time step was set to 2~fs, and the simulations were performed either for 400~ns 
(only counterions), 300~ns (systems with KCl), or 600~ns (systems with CaCl$_2$). The last 
250, 200, or 300~ns of the simulations were included in the analyses, respectively.
